---
title: "Exploratory Data Analysis with R"
author: "Issa Abdulkadir"
date: 03/22/2021
output:
  html_document:
    fig_height: 5
    fig_width: 8
    toc: yes
    toc_float: 
      collapsed: true
    toc_depth: 4  
    
---
***
# Abstract

The purpose of this project is to perform exploratory data analyses on the dataset 
from the peer-to-peer personal loan service provider called Prosper. Since 
a borrower's financial health is based on how many loans the borrower carries 
and the borrower's ability to pay back those loans on time, Prosper uses a 
variety of criteria to approve personal loans- it ranges from credit score, 
income level, and debt-to-income ration, to name just a few. Prosper differs 
from the traditional lenders such as banks or other lending institutions. 

The EDA encompasses mainly the branch of statistics called descriptive 
statistics - mean, mode, median, and standard deviation. This method is used 
to find the central tendency and the spread of our dataset.

This project's EDA process is a stepping stone to learn from the historical data
of personal loans. Analyzing various variables in the dataset and using 
statistical methods to glean insights from the dataset,can shade some light 
on how future borrowers can make better borrowing decisions.


# Overview of Raw Data and Statistical Summarries

The dataset contains 81 variables and 113,937 observations. This is a huge dataset!

In order to make the exploratory analyses steps more digestible, the number of variables has been narrowed down to 23 from 81.
```{r echo=FALSE, message=FALSE, warning=FALSE}
# knitr: Suppress code/messages/warnings 
#        Set default plot options and center them
knitr::opts_chunk$set(fig.width=9,fig.height=5,fig.path='Figs/',
                      fig.align='center',tidy=TRUE,
                      echo=FALSE,warning=FALSE,message=FALSE)
library(dplyr)
library(ggplot2)
prosper_loans <- read.csv('prosperLoanData.csv.')
#prosper_loans <- prosper_loans[c(5,6,8,9,14,15,16,20,22,26,27,29,41:43,45,48,50,54,68,79:81)]
keep_col <- c("Term","LoanStatus","BorrowerAPR","BorrowerRate","ProsperRating..numeric.","ProsperRating..Alpha.", "ProsperScore","EmploymentStatus","IsBorrowerHomeowner","CreditScoreRangeLower","CreditScoreRangeUpper","CurrentCreditLines", "RevolvingCreditBalance", "BankcardUtilization", "AvailableBankcardCredit", "TradesNeverDelinquent..percentage.", "IncomeRange", "StatedMonthlyIncome", "OnTimeProsperPayments", "MonthlyLoanPayment", "InvestmentFromFriendsCount", "InvestmentFromFriendsAmount", "Investors")
prosper_loans <- prosper_loans[keep_col]
class(prosper_loans)
str(prosper_loans)
summary(prosper_loans)

```


# Univariate Variables
Let's start with the univariate data analyses

### Borrowers Rate
```{r echo=FALSE, message=FALSE, warning=FALSE}
# The histogram of borrower rate
library(ggplot2)
ggplot(data = prosper_loans, aes(BorrowerRate)) +
  geom_histogram(binwidth = .007) +
  scale_x_continuous(breaks = seq(0.0,0.5, 0.1) )
```

For investors, knowing the rates to attach to their return on investment is very important in deciding whether to grant loans to potential borrowers. The graph shows the spread of borrowers' rates across many loans.
What is the most common interest rate of borrowers?

Borrowers' rate stats
```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(prosper_loans$BorrowerRate)
```


### Borrowers APR(Annual Percentage Rate)
```{r echo=FALSE, message=FALSE, warning=FALSE}
# The plot of borrower APR is spaced out by sequence on the x axis
ggplot(data = prosper_loans, aes(BorrowerAPR)) +
  geom_histogram(binwidth = .008, aes(y = ..count..)) +
  scale_x_continuous(breaks = seq(0.01,0.5123, 0.05) ) 
  
```

The annual percentage rate(APR) is shown by the graph. The APR plays a crucial role in qualifying for a loan. Borrowers with good spending habits tend to have lower APRs and therefore better in getting loans.

Borrowers' APR stats
```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(prosper_loans$BorrowerAPR)
```

### The Status of the Loans
```{r echo=FALSE, message=FALSE, warning=FALSE}
# The status of the loans showing which the highest group in the category
ggplot(data = prosper_loans, aes(LoanStatus, fill = LoanStatus) ) +
  geom_bar(stat = 'count') +
  coord_flip() +
  theme(legend.position = 'none')
```

The bar graph shows the status of the various loans. Knowing the overall history of the loans status gives insight about the risk and reward associated with this innovative way of lending money.


### Categorical Prosper Rating with NA Omittion 
```{r echo=FALSE, message=FALSE, warning=FALSE}
# The prosperRating is ordered from the lowest to highest count, excluding NA
prosper_loans$ProsperRating..Alpha. <- factor(prosper_loans$ProsperRating..Alpha.,
                                              levels = c('AA', 'A', 'B',
                                                         'C', 'D', 'E', 'HR'),
                                              ordered = TRUE)

ggplot(prosper_loans, aes(x = ProsperRating..Alpha.)) +
  geom_bar(aes(fill = ProsperRating..Alpha.)) +
   xlim('AA', 'A', 'B','C', 'D', 'E', 'HR') +
  theme(legend.position = 'none')
  
```

The graph displays a way Prosper uses to rate its borrowers. The alphabetical rating system takes into considerationg the risk associated with each borrower. For example, AA indicates that a borrower is a low risk and chances of qualifying for a loan is high.

### Prosper Rating in Numbers
```{r echo=FALSE, message=FALSE, warning=FALSE}
# The distribution of prosperRating in numeric arrangment
ggplot(data = prosper_loans, aes(ProsperRating..numeric., 
                                  fill = ProsperRating..numeric.)) +
  
  geom_bar(stat = 'count', binwidth = 0.1, aes(colors = ProsperRating..Alpha.))+ 


  scale_x_continuous(breaks = seq(1,7,1)) +
  theme(legend.position = 'none')

```

The graph displays a way Prosper uses to rate its borrowers. The lower the number of ProsperRating, the higher the risk for lending.

### Loan Term Filled by Employment Status
```{r echo=FALSE, message=FALSE, warning=FALSE}
# Using factor function, the Term variable is adjusted to show an orderly graph.
prosper_loans$Term <- factor(prosper_loans$Term, levels = c(12,36,60),
                              ordered = TRUE)
ggplot(data = prosper_loans, aes(prosper_loans$Term)) +
  geom_bar(aes(fill = prosper_loans$EmploymentStatus)) +
  labs(x = "Loan Terms(Months)",
      fill = "Employment Status")
  
```

Various borrowers have loan terms that range from a year to more than 5 years. Loan terms are most likely influenced by the employment status of the borrowers. As the graph show, most of the loans have terms that are 3 years long.

### Stated Monthly Income Converted to Yearly Income
```{r echo=FALSE, message=FALSE, warning=FALSE}
# The yearlyIcome variable is created to give the incomeRange variable a 
#more orderly format.
prosper_loans$YearlyIncome <- prosper_loans$StatedMonthlyIncome * 12
prosper_loans$IncomeRange <- cut(prosper_loans$YearlyIncome, dig.lab = 10,
                           breaks = c(0, 10000, 20000, 30000, 40000, 50000,
                                      60000, 70000, 80000, 90000, 100000,
                                      200000, 21000036.),
                           labels = c("$0-$10,000", "$10,000-$20,000","$20,000-$30,000",
                  "$30,000-$40,000", "$40,000-$50,000","$50,000-$60,000",
                  "$60,000-$70,000","$70,000-$80,000",
                  "$80,000-$90,000","$90,000-$100,000",
                  "$100,000-$200,000","$200,000-$21000036."))
ggplot(data = prosper_loans, aes(prosper_loans$IncomeRange)) +
  geom_bar(stat = 'count', binwidth = 0.001, 
           aes(fill = prosper_loans$IncomeRange)) +
  coord_flip()  +
  theme(legend.position = "none") 

```

To make easier to understand the reason behind the graph, the monthly income is multiplied by 12 months in a year for each borrower in order to create a yearly income category. Using the transformed graph, the income distribution of the borrowers becomes clear in how their loans amount are granted.

### Current Creditlines
```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(prosper_loans, aes(CurrentCreditLines)) +
  geom_bar()

```

The graphs shows the normal distribution of current creditlines of the Prosper borrowers. A lot of borrowers have similar credit use habit based on the shape of the graph.

### On Time Prosper Payments
```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(prosper_loans, aes(OnTimeProsperPayments)) +
  geom_histogram()
```

The plot shows the number of times borrowers paid their respective loans on time. Interestingly, from the skewness of the data to the left, many borrowers fail to pay their loans in a timely manner.

### Investment From Friends
```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(prosper_loans, aes(InvestmentFromFriendsCount)) +
  geom_bar() +
  xlim(c(1,5)) +
  ylim(c(0,250)) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1),
        text = element_text(size = 20)) 
```

The bar graph show  the frequencies of friends to invest in Prosper loan system. The small number of investment from friends raises a lot of questions and properly warrants its own  further investigation.

Stats About Investment From Friends
```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(prosper_loans$InvestmentFromFriendsCount)
```


### Investors
```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(prosper_loans, aes(Investors)) +
  geom_bar(binwidth = 2 ) +
  xlim(0,700) +
  ylim(0,800)

```

The graph shows the number of Prosper investors. The investors are the bread and butter of the Prosper business model and therefore special attention must be paid to the analysis of this group. The trend shows the number of investors skewing to the left and that signals how a lot of investors are less likely to embrace this type of lending model.

Stats About Investors
```{r}
summary(prosper_loans$Investors)
```
The the number of investors skews to the left as the median and the mean significantly differ.


### Monthly Loan Payments
```{r echo=FALSE, message=FALSE, warning=FALSE}
# The distribution of monthly payments using 
ggplot(prosper_loans, aes(MonthlyLoanPayment)) +
  geom_density() +
  stat_density()

  
```

Exploring the monthly payments of the borrowers shades some light on wich side of the economic spectrum do they fall on.
### Borrowers with Homeownership versus Borrowers Without
```{r}
ggplot(prosper_loans, aes(IsBorrowerHomeowner)) +
  geom_bar(aes(fill = IsBorrowerHomeowner))
```
 
 Homeownership is seen as a symbol of financial responsibility and therefore exploring and comparing the borrowers with homeownership and  borrowers without predict the distribution of how loans are granted.


# Bivariate Analysis

### The Relationship Between Borowers Rate and Borrowers APR
```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = prosper_loans, aes(BorrowerRate, BorrowerAPR)) +
  geom_point(color = '#F79422', alpha = 1/4) +
  stat_smooth(method = 'lm') +
  scale_x_continuous(lim = c(0,quantile(prosper_loans$BorrowerRate, 0.99, na.rm = TRUE))) +
  scale_y_continuous(lim = c(0, quantile(prosper_loans$BorrowerAPR,0.99, na.rm = TRUE))) 
```

 *Is there a correlation between borrowerRate and borrowerAPR?*
From the plot the line that goes through the points seems to disect the data 
into nearly equal halves and most of the points are close to the line. This 
shows a linearly relationship between the rate and the APR. 

The Correlation of Borrowers' Rate and Borrowers' APR in Mathematical Sense 
```{r echo=FALSE, message=FALSE, warning=FALSE}
(with(prosper_loans,cor.test(BorrowerRate,BorrowerAPR)))

```
There's a positive linear relationship between the rate and APR of the borrowers.

*Do the level of income and a debt a borrower has hamper a loan approval?*
Most borrowers that have income less than $25,000 tend to have a high 
debt-to-income ratio that's a red flag in their ability to qualify for 
Prosper loans.


## Correlation of Credit Card Utilization
```{r echo=FALSE, message=FALSE, warning=FALSE}
library(GGally) 
ggpairs(prosper_loans[,c("RevolvingCreditBalance","BankcardUtilization",
                          "AvailableBankcardCredit","CurrentCreditLines")])
```

*Are there any correlations among various variables?*
When plotting how borrowers use their credit cards, there are weak positive and negative correlations.


### Lower Range Credit Scores Stats
```{r}
summary(prosper_loans$CreditScoreRangeLower)
```


### Loan-Term in Relation to Lower-Range Credit Scores Colored by Term
```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = prosper_loans, aes(Term, CreditScoreRangeLower)) +
  geom_boxplot(aes(fill = prosper_loans$Term)) +
  labs(fill = 'Terms in Months')

```

*Is it really true that a borrower must have a minimum of specified credit 
score to qualify for a loan?*

Majority of the borrowers have credit scores greater than 500 which is a big 
indicator that a borrower is financially responsible. There are some outliers 
that indicate that a borrower does not necessary have to meet specified criteria
to be granted a loan. For example, from the graph a borrower or borrowers with 
zero scores were able to successfully received some loans which is an exception in this case(outlier indicator).

### Borrower Rate Versus Lower Range Credit Scores
```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = prosper_loans, aes(BorrowerRate,CreditScoreRangeLower)) +
  geom_jitter(alpha = 1/2, shape = 5, fill = 'yellow') +
  geom_smooth() +
  xlim(0.1,0.5) +
  ylim(400,750) +
  theme(legend.position = "none") +
  theme_dark()
```

When comparing the income of the borrowers' rate versus their credit scores, 
there is a trend that exhibits a higher credit score is associated with lower 
borrower rate and vice versa.
  

# Multivariate Plots

### Current Creditlines and Their Impact on Prosper On Time Payments 
```{r echo=FALSE, message=FALSE, warning=FALSE}
prosper_loans$ProsperScore <- factor(prosper_loans$ProsperScore, levels = c(1,2,3,4,5,6,7,8,9,10), ordered = TRUE)
ggplot(aes(x = CurrentCreditLines, y = OnTimeProsperPayments),
       data = subset(prosper_loans, !is.na(EmploymentStatus))) +
  geom_point(aes(color = ProsperScore), stat = 'summary', fun = median, size = 2) +
  geom_smooth() +
  scale_fill_brewer()
```

Prosper rates its borrowers using ProsperScore ranging from 1-10, with 10 as low risk and 1 as high risk. Using a regression line to describe the relationship between CurrentCreditLines and OnTimeProsperPayments, shows no solid effects as to how one variable affects the other. 

### The Relationship between Amount of Investment from Friends and Their Frequencies
```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = InvestmentFromFriendsAmount, y = InvestmentFromFriendsCount),
       data = subset(prosper_loans, !is.na(IsBorrowerHomeowner))) +
  geom_point(aes(color = Investors), stat = 'summary', fun = mean, size = 2) +
  geom_smooth() +
  xlim(1,15000)+
  scale_color_gradient2(midpoint = mean(subset(prosper_loans, !is.na(IsBorrowerHomeowner))$Investors))
```

The graph shows how much money in USD, friends invested and by how many friends. The relationship is explored because since the idea of the loan is based on peer-to-peer lending, the willingness of friends to invest in their peers is an area of interest.

### Borrowers with Zero Delinquencies
```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = CreditScoreRangeUpper, y = TradesNeverDelinquent..percentage.),
       data = subset(prosper_loans, !is.na(IsBorrowerHomeowner))) +
    geom_point(aes(colour= IsBorrowerHomeowner, shape = IsBorrowerHomeowner), stat = 'summary', fun = mean, size = 2 ) + geom_smooth()
```

*Does owning a home indicate your are a more responsible borrower?*
Through visualizing the relationship between credit score and percentage of 
being never delinquent on any loans, shows that borrowers with homeownership are
likely to be on top of their loans as their percentage of being never delinquent
is way above 50.



### Comparison of Monthly Payments and Stated Monthly Income

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(prosper_loans, aes(MonthlyLoanPayment, StatedMonthlyIncome)) +
  geom_point(fill = I('#F79420'), color = I('black'), shape = 21) +
  geom_smooth() +
  xlim(1,500) +
  ylim(0,5000)
```

*Are borrowers  with monthly income pay more on their monthly payments?*
Apparently, most of the borrowers have payments that range from less than 100 to more than 500 
dollars.
  

# Final Plots and Summary
I found three plots below very compelling in determining how borrowers' are 
granted loans.

### The Income Range
```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = prosper_loans, aes(IncomeRange)) +
  geom_bar(stat = 'count', binwidth = 0.001, aes(fill = IncomeRange)) +
  coord_flip() +
  labs(title = "The Income Range of the Borrowers",
       caption = "Most of the borrowers' income is between $25,000 
       and $74,999",
       x = "The Borrowers' Income Level($)",
       y = "The Number of the Borrowers") +
  theme(
  plot.title = element_text(hjust = 0.5),
  plot.subtitle = element_text(hjust = 0.5),
  plot.caption = element_text(hjust = 0, face = "italic"),
  legend.position  = "none"
  )
  
```

Most borrowers' income accumulate around $25,000 to 
$74,999. This indicates that borrowers are likely to pay their loans back.

### Monthly Payments Stats
```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(prosper_loans$MonthlyLoanPayment)
```

### The Monthly Payments Distribution
```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(prosper_loans, aes(MonthlyLoanPayment)) +
  geom_density() +
  stat_density() +
  labs(x = " Monthly Loan Payments(USD)",
       title = "Distribution of How Much All Borrowers Pay a Month") +
  theme(
  plot.title = element_text(hjust = 0.5),
  plot.subtitle = element_text(hjust = 0.5)
  )
 
```



The monthly payments on the loans are mostly around $200 and with the 
maximum payment around $2000 is an obvious outlier.  
Most of the monthly payments are skewed to the right according to the graph.



### The Value of Homeownership in Relation to Avoiding Delinquencies
```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = CreditScoreRangeUpper, y = TradesNeverDelinquent..percentage.),
       data = subset(prosper_loans, !is.na(IsBorrowerHomeowner))) +
    geom_point(aes(color= IsBorrowerHomeowner, 
                   shape = IsBorrowerHomeowner), 
               stat = 'summary', fun = mean, size = 4 ) + geom_smooth() +
  labs( x = "Upper Credit Scores\n",
        y = "Never Delinquent Loans(%)",
        title = " The Benefit of Homeownership Towards Granting Loans",
        caption = 'Homeownership reflects high credit score')+
  theme(
  plot.title = element_text(hjust = 0.5),
  plot.subtitle = element_text(hjust = 0.5),
  plot.caption = element_text(hjust = 0, face = "italic")
  )
  
  
```


```{r echo=FALSE, message=FALSE, warning=FALSE}
cor.test(prosper_loans$CreditScoreRangeUpper ,prosper_loans$TradesNeverDelinquent..percentage.)
```

Borrowers that own homes have better credit scores and therefore less chances 
of being delinquent on their loans, even though the correlation  between the two variables is not linearly enough to demonstrate a strong relationship. As the credit scores increase, the 
percentage of never delinquent on the loans increases.

# Reflection

### Problems Encountered

Even though the dataset is well compiled with a few errors here and there, 
the missing values posed a very challenging problem in deciding how to relate
various variables in graphical ways. Also, my assumption that certain values 
will fall within expected range was proven wrong as I came across some 
surprising trends in how loans are granted.

Another problem that made me frustrated is my lack of understandinng  in how to
plot a map using the borrowers states. I tried everything from using borrowers'
monthly loan payments and other variables in constructing a visual map of which
states pay more monthly payments on the loans.


### Conclusion
Determing how to grant a loan is a very interesting and multiple-factor process.
When a prospective borrower seeks a loan from a peer-to-peer lender, sometimes 
not all hopes are lost when the borrower's credit history is not sound. Unlike 
the traditional lending institutions, lenders like Prosper take into 
consideration a lot of things but no one factor is proven to be a deciding 
factor in granting a loan.

Homeownership reflects the borrower's ability to maintain above average credit 
score and which in turn helps in speeding a borrowing process from the lender.

A future question related to this dataset is why investors vary from loan to loan and what other metrics need to be included in favor of the borrowers' ability to qualify for loans relatively easily.


# Sources
Black, M. (n.d.). Peer-to-Peer lending: Best websites of March 2021. 
Retrieved March 22, 2021, from 
https://www.investopedia.com/articles/investing/092315/7-best-peertopeer-lending-websites.asp

Remove legend in ggplot2 (3 example codes): Hide one or all legends. (2020, September 30). Retrieved March 24, 2021, from https://statisticsglobe.com/remove-legend-ggplot2-r
























